---
title: "Introduction to R Markdown"
author: "Jaime Undurraga"
date: "`r format(Sys.time(), '%d %B, %Y')`"
# output:
#   html_document: default
#   pdf_document: default
output: ioslides_presentation
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=4, fig.path='Figs/',
                      echo=F, warning=F, message=F)
pandoc_table <- function(df, name, caption, col_p_name="", only.sig=F){
  cnames <- colnames(df)
  b <- sapply(cnames, function(x) x %in% c("p", "p.value", "p-value", "Pr(>F)", col_p_name))
  if (any(b)){
    idx = which(df[, cnames[b]] < 0.05, arr.ind = TRUE)
    if (only.sig){
      df <- df[idx, ]
    }
    else{
      if(length(idx)) emphasize.strong.rows(idx)
    }
  }
  set.caption(caption) 
  pandoc.table(df, split.table = Inf, digits = 1, style="rmarkdown", justify = 'left')
  # tab_nums(name = name, caption = caption)
}

```
## What is R?

- R is a language and environment for statistical computing and graphics. It is a GNU project which is similar to the S language and environment which was developed at Bell Laboratories (much code written for S runs unaltered under R).

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

## why using R and RMarkdown?
1. It is free and open source (GNU General Public License in source code form)
2. Time and processing efficient
3. Many libraries (linear and nonlinear modelling, classical statistical tests, time-series analysis, classification, clustering, etc)
4. Continuosly improved and developed by the R community
5. Well documented (your answer will be usually found in stackoverflow)
6. It makes statistical analysis reproducible to others
7. It is a multiplatform system (FreeBSD, Linux, Windows and MacOS)
8. well-designed publication-quality plots can be produced
9. It is nice

## R IDE
- R Studio is a nice Interface development enviorment to use R 

## Libraries
-There are many R libraries which can be installed and loaded.

## Usefurl links

- [Rmarkdown](http://rmarkdown.rstudio.com/lesson-1.html)

- [Chunks](http://rmarkdown.rstudio.com/authoring_rcodechunks.html)




```{r libraries}
library(xlsx)
```

## Data Frames

- Data Frames contain and allow to manipulate data in many ways. Imported data from Excel, SPPS, cvc, txt and other formats can be loaded into R data frames
```{r}
raw <- read.xlsx('./pta_data.xlsx', 1, encoding = "UTF-8")
head(raw, n=2)

```

## Descriptive statistics

```{r, size = "tiny"}
summary(raw)

```

## Data frame from wide to long format
```{r}
library(reshape2)
raw$subject <- factor(raw$subject)
PTA <- melt(raw, id.vars=c(1:3,16:23), measure.name=4:16, variable.name = "condition",  value.name = "dBHL")
cnd = read.table(text = as.character(PTA$condition), sep = "_", colClasses = "character")
PTA <- cbind(PTA, cnd[c(1,4)])
names(PTA)[names(PTA) == 'V1'] <- 'EAR'
names(PTA)[names(PTA) == 'V4'] <- 'Frequency'
PTA$EAR <- factor(PTA$EAR)
PTA$Frequency = factor(as.numeric(PTA$Frequency))
PTA <- PTA[, -4:-12]
head(PTA, n = 5)
```

## Plotting with ggplot
```{r}
library(ggplot2)
gp <- (ggplot(data = PTA,
        aes(x = Frequency,
            y = dBHL
        ))
+ geom_boxplot(notch=F)
+ geom_jitter(mapping=aes(x=Frequency, y=dBHL), width=0.3, height=0, color='gray')
+ facet_grid(Sex~EAR)
+ xlab("Frequency [Hz]")
+ ylab ("dBHL [dB]"))
```
## Plotting with ggplot
```{r,fig.cap="My cool caption"}
gp
```

## Adding new Factors
```{r}
median_age <- round(median(raw$Age_years))
PTA$AGE_GROUP <- ifelse(PTA$Age_years < median_age, paste("below_", as.character(median_age),sep=""),
                        paste("above_", as.character(median_age),sep=""))
PTA$AGE_GROUP <- factor(PTA$AGE_GROUP)
head(PTA[ ,c(1:5, ncol(PTA))], n=10)
```

## Plotting with ggplot

```{r}
ggplot(data = PTA,
        aes(x = AGE_GROUP,
            y = dBHL,
            color=AGE_GROUP
        )) + 
  geom_violin() + 
  geom_jitter(mapping=aes(x=AGE_GROUP, y=dBHL), width=0.3, height=0, color='gray') + 
  facet_grid(Sex~Frequency) + 
  xlab("Frequency [Hz]") + 
  ylab ("dBHL [dB]")
```

## ANOVA analysis

```{r}
library(ez)
library(pander)
anv = ezANOVA(data = PTA
              , dv = .(dBHL)
              , wid = .(subject)
              , within = .(Frequency, EAR)
              , between = .(Sex, AGE_GROUP)
              , detailed = T
              , type = 2
)
```

## ANOVA analysis

- Tables using ktable 

```{r, results='asis'}
knitr::kable(anv$ANOVA)
```

## 

- Tables using pander

```{r, results='asis'}
pandoc_table(anv$ANOVA, caption="RM - ANOVA")
```  


## ANOVA Mauchly's Test for Sphericity

```{r, echo=FALSE}
row.names(anv$`Mauchly's Test for Sphericity`) <- NULL
```

```{r, results='asis'}
pandoc_table(anv$`Mauchly's Test for Sphericity`, caption="Mauchly's Test for Sphericity")
```  


## ANOVA Sphericity Corrections

```{r, echo=FALSE}
row.names(anv$`Sphericity Corrections`) <- NULL
```


```{r, results='asis'}
pandoc_table(anv$`Sphericity Corrections`, caption="Sphericity Corrections")
```  


## ANOVA plots

```{r}
ezPlot(data = PTA
              , x = Frequency
              , dv = .(dBHL)
              , wid = .(subject)
              , within = .(Frequency, EAR)
              , between = .(Sex, AGE_GROUP)
              , type = 2
              , x_lab = "Frequency [Hz]"
              , y_lab = "dBHL"
              , split = AGE_GROUP
              , col = EAR
              , row = Sex
              , print_code = F)

```


## ANOVA plots

```{r}
ezPlot(data = PTA
              , x = Frequency
              , dv = .(dBHL)
              , wid = .(subject)
              , within = .(Frequency)
              , within_full = .(Frequency, EAR)
              , between_full = .(Sex, AGE_GROUP)
              , type = 2
              , x_lab = "Frequency [Hz]"
              , y_lab = "dBHL"
              , print_code = F)
```

## Correlations
```{r}
library(psych)
library(plyr)
my_corr_f <- function(x,y)
{
  r_corr <- cor.test(x, y)
  b <- lm(x ~ y)
  tab_cor <- data.frame(Value = b$coefficients[2], 
                        Std.Error=summary(b)$sigma, 
                        t = r_corr$statistic, 
                        df = r_corr$parameter, 
                        p = r_corr$p.value, 
                        r = r_corr$estimate)
  row.names(tab_cor) <- NULL
  return(tab_cor)
}
```

## Correlations

```{r}
corrs <- ddply(PTA, .(Sex, EAR, Frequency), function(df) my_corr_f(df$Age_years, df$dBHL))
```

```{r, results='asis'}
pandoc_table(corrs,caption="Cool table")
```  

## Plotting correlations

```{r}
library(ggrepel)
```

## Plotting correlations

```{r}
ggplot(data = PTA,
        aes(x = Age_years,
            y = dBHL,
            color=EAR
        )) + 
         geom_point() + 
         geom_smooth(method="lm", se=T) + 
         geom_text_repel(data=corrs, aes(x=40, 
                                         y=35, 
                                         color=EAR,
                                         label=paste("r=", round(r, digits = 2), "\n p=", round(p, digits = 2), sep=""))) + 
         facet_grid(Sex~Frequency) + 
         xlab("Age [Years]") + 
         ylab ("dBHL [dB]")
```


## Nonparametric tests
- Independent 2-group Mann-Whitney U Test 

```{r}
mean_freq <- ddply(PTA, .(subject, Frequency)
                   , summarise
                   , dBHL = mean(dBHL)
                   , AGE_GROUP = unique(AGE_GROUP))

wt <- wilcox.test(dBHL ~ AGE_GROUP, data=mean_freq) 
tab <- data.frame(W=wt$statistic, p.value=wt$p.value)
```

```{r, results='asis'}
pandoc_table(tab,caption="Nonparametric test")
```  


## Nonparametric tests
- Kruskal Wallis Test One Way Anova by Ranks
```{r}
kt <- kruskal.test(dBHL ~ AGE_GROUP, data=mean_freq) 
tab <- data.frame(chi.squared=kt$statistic, df=kt$parameter ,p.value=kt$p.value)
pandoc.table(tab, split.table = Inf, digits = 1, style="rmarkdown", justify = 'left')

kt <- kruskal.test(dBHL ~ Frequency, data=mean_freq) 
tab <- data.frame(chi.squared=kt$statistic,df=kt$parameter, p.value=kt$p.value)
pandoc.table(tab, split.table = Inf, digits = 1, style="rmarkdown", justify = 'left')

gp <- (ggplot(data = mean_freq,
        aes(x = Frequency,
            y = dBHL
        ))
+ geom_boxplot(notch=T)
+ geom_jitter(mapping=aes(x=Frequency, y=dBHL), width=0.5, height=0, color='gray')
+ xlab("Frequency [Hz]")
+ ylab ("dB HL"))

```

## Nonparametric tests

```{r}
gp
```

## Nonparametric tests
- Friedman Test

```{r}
ft <- friedman.test(dBHL ~ Frequency | subject, data=mean_freq) 
tab <- data.frame(chi.squared=ft$statistic,df=ft$parameter, p.value=ft$p.value)

```

```{r, results='asis'}
pandoc_table(tab,caption="Friedman - Nonparametric test")
```  


